A Method for Regularization of Evolutionary Polynomial Regressions

### 2013-12-19 texts/submission_02-asc/paper_spfga-review.tex
Text revision into subsection 2.2: regularization cost.

\acp{GA} were, arguably, one of the hottest topics of research in the recent decades and with good reason since they outline an optimization scheme easy to conceptualize and with very broad application. If a nonlinear (or otherwize) model requires parameterization, \acp{GA} provide a simple and often effective approach to search for locally optimal parameters. Research related to genetic algorithms abound and spans from the 1950s seminal work of Nils Aall Barricelli \citep{Barricelli:1962aa} in the Institute for Advanced Study of Princeton to today's principal area of study for thousands of researchers, covered in hundreds of conferences, workshops and other meetings. Perhaps the key impulse to \acp{GA} come from John Holland's work and his book ``Adaptation in Natural and Artificial Systems'' \citep{Holland:1975aa}.

@@

Let  $\left\lbrace q_1, \ldots, q_k\right\rbrace$ be a set of monomials over the variables $X_1, \ldots, X_m$. The encoding of that set using $d$ bits per exponent is a binary list such that
\begin{enumerate}
\item the initial segment of $k$ bits defines the active state of each monomial;
\item the remaining bits are split into $k$ segments of size $m\times d$, each representing a monomial;
\item the bits in each monomial segment are split into $m$ sub-segments of size $d$. The $j\nth$ sub-segment is the binary representation of the degree of the variable $X_j$ in the enclosing monomial segment;
\end{enumerate}
This encoding can also be viewed as the flattening of the binary exponents in the matrix representation prefixed by the activation segment. The set $\left\lbrace x_1^3 x_3, x_3^7, x_1 x_2\right\rbrace$ (where $m = 3$) has matrix representation
$$\left[ 
\begin{array}{ccc}
3 & 0 & 1 \\
0 & 0 & 7 \\
1 & 1 & 0
\end{array}
\right] = \left[ 
\begin{array}{ccc}
011 & 000 & 001 \\
000 & 000 & 111 \\
001 & 001 & 000
\end{array}
\right]_{\left( 2 \right)}
$$
where the right matrix is in binary form with  $d = 3$ bits. An encoding where the extra monomial $x_1^6x_2^2x_3^5$ is inactive, setting $k = 4$, would be
\begin{center}
1110;~011, 000, 001;~000,000,111;~001,001,000;~110,010,101
\end{center}
where, for reading purposes, semicolons separate segments and commas separate variables. The first $k=4$ bits inform that the first, second and third monomials are active while the fourth is not.

While each valid encoding represents a set of monomials the map is not bijective: each set of monomials has multiple encodings, for example by changing $d$ or the order of monomial segments. However, considering the \ac{EPR} task, this is a minor issue and a bijective map would bring computational complexity and impact negatively the algorithm's performance.

There is one final remark concerning the encoding. As it is, the activation segment can become all zeros, representing the empty set of monomials. This situation can be avoided with a simple hack: Given an encoding, the first monomial is always considered active, thus restricting the syntactic form of encodings to binary strings starting with $1$. In practice, this means that the implementation of the encoding can omit the first bit.
% If the coding consists of entirely zeros, by convention, it describes polynomial $x_1$. This has to do with the execution of the polynomial regression that would fail if we interpret it as the zero polynomial. Anyway, for progressive larger  binary descriptions, the chances of getting this zero description decrease exponentially, so it does not impact in any meaningful way in the algorithm's performance.

@@

The polynomial regression \ac{rmse} considered so far is based on the ability to predict the test set after the polynomial regression has found the appropriate coefficients $\theta_i$ for each one of the monomials $q_i$.

A known problem of using a cost function based only in the dataset \ac{rmse} and with polynomial regressions in general is a tendency to overfit the training data. The excessive variance is can be reduced by a regularization of the \ac{rmse} function. Tnd the inclusion of a penalty factor proportional to the number of monomials is a possible regularization method. Thus, the error from equation \ref{eq:rmse} defines
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta;\mathcal{Q},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{Q},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized.

@@

\caption{Error distribution by regularization exponent for the Abalone and Auto MPG datasets. The box plots summarize the error values of 10 runs for each value of $\lambda$. The smallest overall error is shown in grey and, in both cases, was achieved with $\lambda=0.8$.}
\label{Abalone_dataset_lambdas}\label{Auto-MPG_dataset_lambdas}

@@

The datasets used herein were selected from Luís Torgo's data  repository, \url{http://www.dcc.fc.up.pt/~ltorgo/Regression/DataSets.html}. Most come originally from UCI ML repository, \url{http://archive.ics.uci.edu/ml/}.

@@

%\section{Bibliography}

>>>>

\acp{GA} were, arguably, one of the hottest topics of research in the recent decades and with good reason since they outline an optimization scheme easy to conceptualize and with very broad application. If a nonlinear (or otherwise) model requires parameterization, \acp{GA} provide a simple and often effective approach to search for locally optimal parameters. Research related to genetic algorithms abound and spans from the 1950s seminal work of Nils Aall Barricelli \citep{Barricelli:1962aa} in the Institute for Advanced Study of Princeton to today's principal area of study for thousands of researchers, covered in hundreds of conferences, workshops and other meetings. Perhaps the key impulse to \acp{GA} come from John Holland's work and his book ``Adaptation in Natural and Artificial Systems'' \citep{Holland:1975aa}.

@@

Let  $\left\lbrace q_1, \ldots, q_k\right\rbrace$ be a set of monomials over the variables $X_1, \ldots, X_m$. The encoding of that set using $d$ bits per exponent is a binary list such that
\begin{enumerate}
\item the initial segment of $k$ bits defines the active state of each monomial;
\item the remaining bits are split into $k$ segments of size $m\times d$, each representing a monomial;
\item the bits in each monomial segment are split into $m$ sub-segments of size $d$. The $j\nth$ sub-segment is the binary representation of the degree of the variable $X_j$ in the enclosing monomial segment;
\end{enumerate}
This encoding can also be viewed as the flattening of the binary exponents in the matrix representation prefixed by the activation segment. The set $\left\lbrace x_1^3 x_3, x_3^7, x_1 x_2\right\rbrace$ (then $m = 3$) has matrix representation
$$\left[ 
\begin{array}{ccc}
3 & 0 & 1 \\
0 & 0 & 7 \\
1 & 1 & 0
\end{array}
\right] = \left[ 
\begin{array}{ccc}
011 & 000 & 001 \\
000 & 000 & 111 \\
001 & 001 & 000
\end{array}
\right]_{\left( 2 \right)}
$$
where the right matrix is in binary form with  $d = 3$ bits. An encoding of this set of monomials with the extra monomial $x_1^6x_2^2x_3^5$ inactive, setting $k = 4$, would be
\begin{center}
1110;~011, 000, 001;~000,000,111;~001,001,000;~110,010,101
\end{center}
where, for reading purposes, semicolons separate segments and commas separate variables. The first $k=4$ bits inform that the first, second and third monomials are active while the fourth is not.

While each valid encoding represents a set of monomials the map is not bijective: each set of monomials has multiple encodings, for example by changing $d$ or the order of monomial segments. However, considering the \ac{EPR} task, this is a minor issue and a bijective map would add computational complexity and negative impact to the algorithm's performance.

There is one final remark concerning this encoding method. As it is, the activation segment can become all zeros, representing the empty set of monomials. This situation can be avoided with a simple hack: Given an encoding, the first monomial is always considered active, thus restricting the syntactic form of encodings to binary strings starting with $1$. In practice, this means that the implementation of the encoding can omit the first bit.
% If the coding consists of entirely zeros, by convention, it describes polynomial $x_1$. This has to do with the execution of the polynomial regression that would fail if we interpret it as the zero polynomial. Anyway, for progressive larger  binary descriptions, the chances of getting this zero description decrease exponentially, so it does not impact in any meaningful way in the algorithm's performance.

@@

The polynomial regression \ac{rmse} considered so far accounts for the ability to predict the transformed testset. A known problem of using a cost function based only in the dataset \ac{rmse} (and of polynomial regressions in general) is a tendency to overfit the training data. This excessive variance can be reduced by regularizing the \ac{rmse} function and one possible method is the inclusion of a penalty factor proportional to the number of monomials. Thus, the error from equation \ref{eq:rmse} defines the regularized version
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta;\mathcal{Q},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{Q},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized.

@@

\caption{Error distribution by regularization exponent for two common datasets. The box plots summarize error values of ten simulations for each value of $\lambda$. The smallest overall error, in grey, is achieved in both datasets when $\lambda=0.8$.}
\label{Abalone_dataset_lambdas}\label{Auto-MPG_dataset_lambdas}

@@

Datasets used herein were selected from Luís Torgo's data repository, \url{http://www.dcc.fc.up.pt/~ltorgo/Regression/DataSets.html}. Most can also be found in the UCI ML repository at \url{http://archive.ics.uci.edu/ml/}.

@@

\section*{Bibliography}
